---
title: "Test Mean Squared Error (MSE) Results For All Tournament Sizes and Dynamic Epsilon Lexicase Selection"
author: "Jose Guadalupe Hernandez"
date: "`r Sys.Date()`"
output: bookdown::pdf_book
documentclass: article
description: "Supplementary Material"
geometry: margin=1.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  fig.align = 'center'
)
```

```{r load-libraries, include=FALSE}
rm(list = ls())
library(ggplot2)
library(cowplot)
library(dplyr)
library(PupillometryR)
library(ggpubr)
library(knitr)
library(tidyr)
```

```{r theme-setup, include=FALSE}
# Define conditions and visual parameters
SELECTION <- c('Lexicase', 'T2', 'T5', 'T10', 'T25', 'T50', 'T100', 'T250')
SHAPE <- c(21, 21, 21, 21, 21, 21, 21, 21, 21)
cb_palette <- c('#000000', '#0072B2', '#E69F00', '#009E73', '#D55E00', '#CC79A7', '#56B4E9', '#F0E442', '#999999')
TSIZE <- 17

# Theme for plots
 p_theme <- theme(
  plot.title = element_text(face = "bold", hjust=0.5),
  panel.border = element_blank(),
  panel.grid.minor = element_blank(),
  axis.title = element_text(size=17),
  axis.text = element_text(size=11),
  axis.text.y = element_text(angle = 90, hjust = 0.5),
  axis.title.y = element_blank(),
  legend.position = "none",
  panel.background = element_rect(fill = "#f1f2f5",
                                  colour = "white",
                                  size = 0.5, linetype = "solid")
)
```

```{r load-data, include=FALSE}
# Load the Parent Selection Generator data
all_data <- read.csv("all_tournament_sizes.csv")
all_data$Selection <- factor(all_data$Selection, levels = SELECTION)
```

```{r helper-functions, include=FALSE}
summarise_data <- function(data) {
  data %>%
    group_by(Selection) %>%
    summarise(
      count = n(),
      min = min(performance, na.rm = TRUE),
      p25 = quantile(performance, 0.25, na.rm = TRUE),
      median = median(performance, na.rm = TRUE),
      mean = mean(performance, na.rm = TRUE),
      p75 = quantile(performance, 0.75, na.rm = TRUE),
      max = max(performance, na.rm = TRUE)
    )
}

plot_data <- function(data){
  ggplot(data, aes(x = Selection, y = performance, color = Selection, fill = Selection, shape = Selection)) +
  geom_flat_violin(position = position_nudge(x = .1, y = 0), scale = 'width', alpha = 0.2, width = 1.5) +
  geom_boxplot(color = 'black', width = .07, outlier.shape = NA, alpha = 0.0, size = 1.0, position = position_nudge(x = .16, y = 0)) +
  geom_point(position = position_jitter(width = 0.02, height = 0.0001), size = 1.5, alpha = 1.0) +
  scale_y_log10(name = "Mean Squared Error") +
  scale_x_discrete(name = "Parent Selection Generator") +
  scale_shape_manual(values = SHAPE) +
  scale_colour_manual(values = cb_palette) +
  scale_fill_manual(values = cb_palette) +
  ggtitle('Mean Squared Error on Testing Set') +
   p_theme + coord_flip()
}

# Friedman test for the paired test design
friedman_test <- function(data) {
  friedman.test(performance ~ Selection | Rep, data = data)
}

# Wilcoxon rank-sum test for pairwise comparisons with Holm correction
pairwise_wilcox_test <- function(data) {
  # pivot data to wide format for paired test design
  long_aligned <- data %>%
  mutate(Rep = as.integer(Rep),
         Selection = factor(Selection)) %>%
  arrange(Selection, Rep)

  # print(long_aligned)
  pairwise.wilcox.test(long_aligned$performance, long_aligned$Selection, p.adjust.method = "holm", alternative = "t", paired = TRUE)
}

```

# Background

Compared how Dynamic Epsilon Lexicase Selection compairs against Tournament Selection with different tournament sizes.
Implemented a DEAP evolutionary setup, with a population size of 250 evolved for 100 generations.
Conducted a paired study, where a given block of lexicase and tournament selection runs are aligned by their random seed and used the same data splits (train, validate, test).
We evaluate the performance of each selection strategy on six regression problems: Airfoil, Concrete, Energy Cold, Energy Heat, Housing, and Yacht. Each problem is evaluated using the Mean Squared Error (MSE) on a held-out testing set.
The results are visualized using violin plots to show the distribution of MSE values for each selection strategy across 41 independent runs.
Tasks and experimental setup were obtained from the paper `Down-Sampled Epsilon-Lexicase Selection for Real-World Symbolic Regression Problems` by Geiger, Sobania, and Franz (2023). Link [here](https://dl.acm.org/doi/abs/10.1145/3583131.3590400).
Population size and generations were adjusted to 250 and 100, respectively, to keep the number evaluations the same as the original study and to explore how population size impacts selection.

## Statistical Analysis

We employ two statistical tests:

1. **Friedman Test**: A non-parametric test to detect differences across the nine selection strategies for each task. If p<0.05, we reject the null hypothesis of equal distributions across conditions.
2. **Pairwise Wilcoxon Signed Rank test**: Post-hoc tests with Holm correction to test for two-tailed differences between selection strategies.

# Airfoil

```{r air-data, include=FALSE}
# Load the Parent Selection Generator data
problem_data <- filter(all_data, Task == 'Airfoil')
```

```{r air-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```


```{r air-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Concrete

```{r concrete-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Concrete')
```

```{r concrete-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r concrete-summary, echo=FALSE}
problem_data %>%summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Crime

```{r crime-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Crime')
```

```{r crime-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r crime-summary, echo=FALSE}
problem_data %>%summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Energy Cold

```{r ec-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Energy-C')
```

```{r ec-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r ec-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Energy Heat

```{r eh-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Energy-H')
```

```{r eh-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r eh-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Housing

```{r house-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Housing')
```

```{r house-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r house-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Parkinsons Motor

```{r pkm-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Parkinsons-M')
```

```{r pkm-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r pkm-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Parkinsons Total

```{r pkt-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Parkinsons-T')
```

```{r pkt-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r pkt-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Wine Red

```{r wine-red-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Wine-R')
```

```{r wine-red-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r wine-red-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Wine White

```{r wine-white-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Wine-W')
```

```{r wine-white-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r wine-white-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```

# Yacht

```{r yacht-data, include=FALSE}
problem_data <- filter(all_data, Task == 'Yacht')
```

```{r yacht-plot, out.width="100%"}
problem_data$Selection <- factor(problem_data$Selection, levels = SELECTION)
problem_data %>% plot_data()
```

```{r yacht-summary, echo=FALSE}
problem_data %>% summarise_data()

problem_data %>% friedman_test()

problem_data %>% pairwise_wilcox_test()
```
